---
title: "2020 NFL"
author: "      "
date: "2024-06-12"
output: 
  word_document: 
    fig_caption: yes
---

# EDA OF 2020 NFL PREDICTIONS

In this project, we shall clean the 2020 nfl dataset and perform some simple Exploratory Data Analyis. At the end, we shall export the cleaned data to excel

## Import libraries
```{r}
#We first import the required libaries
library(ggplot2)
library(tidyverse)
library(skimr)
library(broom)
library(Hmisc)
library(readxl)
library(readr)
library(dplyr)
```

## Import dataset and Convert to Desirabe datatypes

```{r}
#Importing the data
#Importing sheet 1
sheet1<-read_excel("C:\\Users\\KEVINS\\OneDrive\\Desktop\\TAB\\nfl_elo_latest.xlsx",sheet = "Sheet 1")
#Importing sheet 2
sheet2<-read_excel("C:\\Users\\KEVINS\\OneDrive\\Desktop\\TAB\\nfl_elo_latest.xlsx",sheet = "Sheet2")
```

```{r}
#We skim through sheet1 and sheet2
#Both the sheets have the same number of columns,30
skim(sheet1)
skim(sheet2)
#Check the data types
str(sheet1)
str(sheet2)
```

```{r}
#We check if the two sheets have identical column names before combining them
identical(names(sheet1),names(sheet2))
#Converting various variables to numeric
sheet1$elo1_pre<-as.numeric(sheet1$elo1_pre)
sheet2$elo1_pre<-as.numeric(sheet2$elo1_pre)
sheet1$elo2_post<-as.numeric(sheet1$elo2_post)
sheet2$elo2_post<-as.numeric(sheet2$elo2_post)
sheet1$qb1_value_pre<-as.numeric(sheet1$qb1_value_pre)
sheet2$qb1_value_pre<-as.numeric(sheet2$qb1_value_pre)
sheet1$score1<-as.numeric(sheet1$score1)
sheet2$score1<-as.numeric(sheet2$score1)
sheet1$score2<-as.numeric(sheet1$score2)
sheet2$score2<-as.numeric(sheet2$score2)
sheet1$qb2_value_pre<-as.numeric(sheet1$qb2_value_pre)
sheet2$qb2_value_pre<-as.numeric(sheet2$qb2_value_pre)
sheet1$qb1_adj<-as.numeric(sheet1$qb1_adj)
sheet2$qb1_adj<-as.numeric(sheet2$qb1_adj)
#Now combine
data_2020nfl<-bind_rows(sheet1,sheet2)
view(data_2020nfl)
```

```{r}
#Using our new data now, we now check the data types
str(data_2020nfl)
#Our data types are perfectly ok
print(data_2020nfl)
```

## Data Cleaning

### Remove columns that are not useful

```{r}
#Columns like date, season, neutral are not important. so we will drop them.
#Also, the playoff column contain many NAs making it irrelevant. We shall also remove it.
data_2020nfl<-data_2020nfl%>%select(-date,-season,-neutral,-playoff)
str(data_2020nfl)
```

## Dealing with Nulls

```{r}
#To deal with Nulls, we shall calculate the columns means and replace the NAs with the mean
#Calculating column means

means_data<-data_2020nfl%>%select(-team1,-team2,-qb1,-qb2) #Selecting columns that have numeric variables
str(means_data)
char_data<- data_2020nfl%>%select(team1,team2,qb1,qb2)#Selecting character variables to be rejoined with the numerical variables after performing mean imputation
means_col<-colMeans(means_data,na.rm = TRUE) #Obtaining the mean of all numeric variables
print(means_col) #Printing the column means.
```

```{r}
#Perform mean imputation
Mean_impute<-function(x){
  col_mean<-means_col
  x[is.na(x)]<-col_mean
  return(x)
}
#Apply the function to each column
means_data[]<-lapply(means_data,Mean_impute)
skim(means_data) #Checking if there are any missing values
#Recombine our data
#Our character variables have missing values. Instead of dropping them, we shall replace them with a value NONE.
char_data[is.na(char_data)]<-"NONE"
skim(char_data)
nfl_2020<-bind_cols(char_data,means_data)

```


```{r}
#Checking if nulls are dealt with
skim(nfl_2020)#There are no more nulls

```

## Dealing with Outliers

```{r}
#We check for outliers in our numerical variables by creating boxplots
means_data<-means_data%>%
  pivot_longer(cols = everything(), names_to = "variable",values_to = "value")
ggplot(means_data,aes(x=variable,y=value))+
  geom_boxplot()+
  ggtitle("Boxplot to identify outliers")
```

```{r}
#From the boxplot, there are outliers.
#We shall cap them using the function below
outlier_cap<-function(x, lower_quantile=0.5, upper_quantile=0.95){
  lower_bound<-quantile(x,lower_quantile)
  upper_bound<-quantile(x,upper_quantile)
  x[x<lower_bound]<-lower_bound
  x[x>upper_bound]<-upper_bound
  return(x)
}
#Apply the fucnction
nfl_2020<-nfl_2020%>%
  mutate(across(c(5:26),~outlier_cap(.)))
print(nfl_2020)

```

```{r}
#Exploring the data
#Creating histogram of different variables
ggplot(nfl_2020,aes(x=elo1_pre))+
  geom_histogram(binwidth = 10,fill="blue",color="black",alpha=0.7)+
  labs(title = "Histogram of elo1_pre",x="Values in elo1_pre",y="Frequency")+
  theme_minimal()

```

```{r}
#adding a histogram of score 1 
ggplot(nfl_2020,aes(x=score1))+
  geom_histogram(binwidth = 10,fill="green",color="black",alpha=0.7)+
  labs(title = "Histogram of score1",x="Values in score1",y="Frequency")+
  theme_minimal()
#Adding a histogram of score 2
ggplot(nfl_2020,aes(x=score1))+
  geom_histogram(binwidth = 10,fill="yellow",color="black",alpha=0.7)+
  labs(title = "Histogram of score2",x="Values in score2",y="Frequency")+
  theme_minimal()
```

```{r}
#Creating density plot to visualize the density distribution of score1 and score2
ggplot(nfl_2020,aes(x=score1))+
  geom_density(fill="blue",alpha=0.6)+
  labs(title="Density plot of Score1",x="Values in Score1")+
  theme_minimal()
ggplot(nfl_2020,aes(x=score2))+
  geom_density(fill="brown",alpha=0.6)+
  labs(title="Density plot of Score2",x="Values in Score2")+
  theme_minimal()

```


## Exporting our cleaned data to Excel

```{r}
#Export the data to excel
writexl::write_xlsx(nfl_2020,"C:\\Users\\KEVINS\\OneDrive\\Desktop\\TAB\\NFL2020.xlsx")

```







